---
title: "Readme"
format: html
editor: visual
---

# Running the project

This project uses targets, targets and tarchetypes so ensure you have the requirements installed already. Run the makefile before you do anything here.

-   Check my notes

# Initial EDA

This was all part of Karl's requests

No data was missing and we've cleaned up naming and stuff:

```{r}
pacman::p_load(tidyverse, targets, lme4, lmerTest, mixedup, gt, skimr)
tar_load(cleaned_data)
skim(cleaned_data)
```

Let's look at the response:

![](figs/gene.svg)

We can see that it's quite right skewed, so a transform might be needed.

Now lets look at some combos, Karl wants to see if the treatment has an effect on gene expression and there's a bunch of other potential dependent variables:

Most aren't interesting or show no clear relationships (Celltype and cellline are related which makes things weirder) but here's a few that are good:

![](figs/geneVsConcByName.svg)

This shows that there are potential individual effects per name!

![](figs/geneVsConcByCell.svg)

This shows that it's not just the cell line (As there's no pattern), but the actual type!![](figs/geneVsConcByTreament.svg)

Here we see that it appears that treatment alone does not have a significant influence on gene expression with concentration.

![](figs/geneVsname.svg)

Name and gene expression alone certainly have individual effects

![](figs/geneVsTreament.svg)

We see here that there is overlap in the IQR between the two which does indicate there might not be a significant difference alone.![](figs/geneVsTreatmentByCell.svg)

We see that treatment and cell-line have a strange interaction, wild-type loves the treatment but cell-type 101 doesn't care

Karl made us create this plot and it's actually the best at visualising stuff:

![](figs/interactionPlot.svg)

That there is defs individual effects coming from the name of the cell-line. Additionally for many of the for the wild-type, there is a clear relationship between treatment and gene-expression which shows there might be a three-way interaction.

# Modelling

We see from that deffs individual effects as well as lots of possible interaction (including three-way interaction) and let's use see if a mixed-effects model works since experimentally it makes sense. We'll start with a simple no interaction model and use ranova to see if the random intercept is useful:

```{r}
base_model <- lme4::lmer(gene_expression ~ conc+cell_line+treatment+(1|name), data = cleaned_data)
ranova(model = base_model) 
```

We can see from this that the random intercept is important, there are individual effects.

So let's start with a threeway interaction model with a random intercept for name and see if it can be reduced using step! Importantly, we need to make the difficult decisions around lower order effects if they're not significant but the higher order ones are. Due to the principle of marginality and the potential to induce bias then I'll be likely keeping non-significant terms in.

Again, we use step to find the best model from our threeway interaction, by default it's backwards so it's removing terms

```{r}
tar_load(threeway_model)
mixedup::summarise_model(threeway_model)
```

As seen in here, we keep the 3-way interaction coefficient. However I'm worried about the insignificance of the lower order terms. Let's do a model without the 3-way interaction, again using backwards steps to remove coefficients:

```{r}
tar_load(twoway_model)
mixedup::summarise_model(twoway_model)
```

Here treatment is not significant. While I doubt it's actually better, let's try a no interaction model.

```{r}
tar_load(simple_model)
mixedup::summarise_model(simple_model)
```

This has actually removed treatment fully, leaving only concentration. This doesn't seem correct since we clearly saw the interaction was significant.

Let's compare the three using AIC/ANOVA:

```{r}
anova(simple_model, twoway_model, threeway_model, refit=FALSE)
tar_load(AIC_table)
AIC_table
```

AIC wants the three-way interaction model and the anova rules out the simple model (as we expected!) while also picking the three-way over the two-way.

So let's stick with the three way, it's AIC is quite a bit lower although explaining the interaction between treatment, concentration and cell_line will be annoying (the two way model was much simpler as it didn't even include the cell_line! Although the plot does show that there is some effect from this.

# Final model

```{r}
tar_load(coef_table)
coef_table
```

We need to check the assumptions so lets also do that:

![](figs/assumptions_model.svg)

The reiduals vs fitted are not great

The QQ is good for the middle values which might be the issue

The scale location plot is confusing, but

## Sample size for new data/model

Karl asked us for a sample size calculation for some new data and a new model for a publication:

```{r}
tar_load(new_model_sample_size)
new_model_sample_size
```

# Appendix

```{r}
here::here()
```
